Basic concepts and tools for the Toki Pona minimalist and constructed language

### 2018-07-01 article/acm/samplebody-journals.tex
terminado historico

\subsection{Historical note}
TP was developed as an internal and personal language
by Sonja Lang~\cite{interview}.
It was released as a draft in 2001 and in 2007 some documents
reported it to have a few hundred speakers.
The English official book~\cite{tpLang} was released only in 2014.
In 2016, a version of the official book was released in French.
Nowadays, one finds a number of texts about TP and written
in it
(e.g. in social platforms such as Facebook groups,
microblogging, Telegram and
IRC), and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.

\subsection{Natural and constructed and artificial languages}
A language one uses (or might use) to communicate by speaking
and writing is called `natural language'.
A `constructed language' (also planned or invented language) is a natural language built by someone or
a group, such as Esperanto, Toki Pona, and Lojban.
An artificial language is a language yield by artificial agents,
such as by AI routines, or by humans in controlled experiments,
and are considered e.g. within `cultural evolution' studies.
Formal languages are defined by tokens and rules to
operate them, they span from computer programming languages
to math and formal models for natural languages.

Constructed languages are in some traditions called artificial
languages, but creators most often prefer to use the term
`standardized', 'constructed'
or 'planned' language with the argument that the conlang
is rooted on natural languages,
and 'artificial' is misleading.
The preferred terms seem to be planned or constructed language
or simply conlang.
The construction of languages is called glossopoeia.
TP is a conlang which might be classified
as engineered for experiments, meditation and philosophy;
and suitable for use as an auxiliary international language and
as an artistic language.~\cite{conlanWikip,artlanWikip}
% taxonomy, vocabulary, ontology, terminology,

>>>>

\subsection{Historical note}
TP was developed as an internal and personal language
by Sonja Lang in late 1990s~\cite{interview}.
It was released as a draft in 2001, and in 2007 some documents
reported it to have a few hundred speakers.
The English official book~\cite{tpLang} was released only in 2014.
In 2016, a version of the official book was released in French.
Nowadays, one finds a number of texts about TP and written
in TP (e.g. in social platforms such as Facebook groups,
microblogging, Telegram and IRC),
and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.

\subsection{Natural and constructed and artificial languages}
A language one uses (or might use) to communicate by speaking
and writing is called `natural language'.
A `constructed language' (also planned or invented language) is a natural language built by someone or
a group, such as Esperanto, Toki Pona, and Lojban.
An `artificial language' is most usually a term used
to designate a language yield by artificial agents,
such as by AI routines, or by humans in controlled experiments,
and are considered e.g. within `cultural evolution' studies.
Formal languages are defined by tokens and rules to
operate them, they span from computer programming languages
to math and formal models for natural languages.

Constructed languages (aka. conlangs) are in some traditions called artificial
languages, but creators most often prefer to use the term
`standardized', 'constructed'
or 'planned' language with the argument that the conlangs
are rooted on natural languages,
and 'artificial' is thus misleading.
The preferred terms seem to be \emph{planned} or \emph{constructed} language
or simply \emph{conlang}.
The construction of languages is called glossopoeia.
TP is might be considered
engineered for experiments, meditation and philosophy;
and suitable for use as an auxiliary international language and
as an artistic language.~\cite{conlanWikip,artlanWikip}
% taxonomy, vocabulary, ontology, terminology,
